---
title: "Building blocks of tidytargets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building blocks of tidytargets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.output = "r"
)
library(tidytargets)

work <- tempfile("tidytargets-vignette-")
dir.create(work)
knitr::opts_knit$set(root.dir = work)
```

tidytargets is a tidy, pipe-friendly grammar for `{targets}`. 

`tt_initialise()` constructs a `tidytargets` object. Each method is a [target factory](https://books.ropensci.org/targets/static.html#target-factories): it appends `tar_target` definitions to a pipeline script. You compose the pipeline with `|>`; `{targets}` runs the dependency graph.

The composition is **lazy**. `tt_initialise()`, `tt_single()`, `tt_iterate()`, `tt_merge()`, and `tt_report()` only write factories into `{store}.R`. User functions are not called until `tt_evaluate()` (or evaluating the object; e.g., printing it in the console), which closes the script and runs `targets::tar_make()`.
Assigning the object does not print it, so an interactive session then says the pipeline is ready to be evaluated, rather than appearing to do nothing.

Pipelines are **incremental**. Add a step, change a parameter, or grow the input list and `{targets}` skips up-to-date targets, re-running only what the new graph requires. That is why this vignette can show the script after each building block without executing the pipeline.

This vignette walks through every building block and prints the targets script after each step with `show_targets_script()`.

## Building blocks

| Function | Role |
| --- | --- |
| `tt_initialise()` | Start a pipeline: write the script header, register inputs |
| `tt_single()` | Add one non-iterated target |
| `tt_iterate()` | Map a function over inputs |
| `tt_merge()` | Combine iterated results into one object |
| `tt_report()` | Render a Quarto / R Markdown report |
| `tt_evaluate()` | Close the script and run `tar_make()` |
| `tt_explore()` | Return one stored instance of a named target |
| `tt_factory()` | Internal factory used by iterate / single / merge |

`tt_factory()` is not meant to be called directly. `tt_single()`, `tt_iterate()`, and `tt_merge()` all append `target_list <- target_list |> target_append(tt_factory(...))` to the script.

## Toy inputs

`tt_initialise()` takes a named list of in-memory objects, one element per unit of iteration (e.g. sample). A named vector of file paths works the same way.

```{r inputs}
inputs <- list(
  sample_a = 1:3,
  sample_b = 4:6
)
```

## `tt_initialise()`

This writes `{store}.R`, saves the inputs, and registers two mapped targets:

- `input_list` — the named input vector or list (override with `target_output`)
- `sample_names` — the names of that vector or list

With no `computing_resources`, the pipeline is sequential. With no `store`, a unique `./tidytargets-<HASH>` directory is created.

```{r initialise}
pipe <- inputs |>
  tt_initialise()
```

```{r initialise-script, comment = ""}
show_targets_script(pipe)
```

## `tt_single()`

This command add an operation on a whole target (not iterative/grouped)

```{r single}
pipe <- pipe |>
  tt_single(
    target_output = "n_samples",
    command = length(sample_names)
  )
```

```{r single-script, comment = ""}
show_targets_script(pipe)
```

`tt_initialise()` itself uses `tt_single()` for the `*_file` targets (`format = "file"`).

## `tt_iterate()`

Map a function over every input. Write `command` as you would in `tar_target()`.

```{r iterate-read}
pipe <- pipe |>
  tt_iterate(
    target_output = "data",
    command = input_list * 2
  )
```

```{r iterate-read-script, comment = ""}
show_targets_script(pipe)
```

A second iterate maps over the result of the first:

```{r iterate-length}
pipe <- pipe |>
  tt_iterate(
    target_output = "n",
    command = length(data)
  )
```

```{r iterate-length-script, comment = ""}
show_targets_script(pipe)
```

## `tt_evaluate()`

```{r evaluate}
pipe |> tt_evaluate()
```

## `tt_explore()`

Return one stored instance of a target so you can inspect it or pipe it onward. A heading names the target (and which instance, for collections). Mapped targets load a single branch rather than every sample.

```{r tt-explore}
pipe |> tt_explore("data")
```

## `tt_merge()`

Collect every branch of an iterated target into one object.

```{r merge}
pipe <- pipe |>
  tt_merge(
    target_output = "n_total",
    command = sum(unlist(n))
  )
```

```{r merge-script, comment = ""}
show_targets_script(pipe)
```

## `tt_report()`

Appends a `tarchetypes::tar_quarto_raw()` factory. Rendering happens later, when you evaluate the pipeline; the grammar only writes the target.

```{r report-file}
writeLines(
  c(
    "---",
    "title: Example report",
    "format: html",
    "params:",
    "  n_total: 0",
    "  n_samples: 0",
    "---",
    "",
    "Samples: `r params$n_samples`. Values: `r params$n_total`."
  ),
  "example-report.qmd"
)
```

```{r report}
pipe <- pipe |>
  tt_report(
    target_output = "report",
    rmd_path = "example-report.qmd",
    params = list(n_total = n_total, n_samples = n_samples)
  )
```

```{r report-script, comment = ""}
show_targets_script(pipe)
```

## Custom functions from a file

`tt_iterate()`, `tt_single()`, and `tt_merge()` accept `user_function_source_path`. The path is written as `source(...)` in the targets script so workers can load helpers that are not in a package.

```{r source-path, eval = FALSE}
pipe |>
  tt_iterate(
    target_output = "transformed",
    command = my_transform(data),
    user_function_source_path = "helpers.R"
  )
```
