---
title: "Building blocks of tidytargets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building blocks of tidytargets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.output = "r"
)
library(tidytargets)

work <- tempfile("tidytargets-vignette-")
dir.create(work)
knitr::opts_knit$set(root.dir = work)

show_script <- function(store) {
  path <- paste0(store, ".R")
  cat("## ", path, "\n\n", sep = "")
  writeLines(readLines(path))
}
```

tidytargets is a tidy, pipe-friendly grammar for `{targets}`. Each grammar function is a [target factory](https://books.ropensci.org/targets/static.html#target-factories): it appends `tar_target` definitions to a pipeline script. You compose the pipeline with `|>`; `{targets}` runs the dependency graph.

This vignette walks through every building block and prints the targets script after each step.

## Building blocks

| Function | Role |
| --- | --- |
| `hpc_initialise()` | Start a pipeline: write the script header, register inputs |
| `is_target()` | Mark an argument as an upstream target name |
| `hpc_iterate()` | Map a function over inputs (and optional tiers) |
| `hpc_single()` | Add one non-iterated target |
| `hpc_merge()` | Combine iterated results into one object |
| `hpc_report()` | Render a Quarto / R Markdown report |
| `hpc_evaluate()` | Close the script and run `tar_make()` |
| `hpc_factory()` | Internal factory used by iterate / single / merge |

`hpc_factory()` is not meant to be called directly. `hpc_iterate()`, `hpc_single()`, and `hpc_merge()` all append `target_list |> target_append(hpc_factory(...))` to the script.

## Toy inputs

`hpc_initialise()` takes a named vector of inputs, typically file paths. Here two RDS files stand in for samples.

```{r inputs}
files <- c(
  sample_a = "a.rds",
  sample_b = "b.rds"
)
saveRDS(1:3, files[["sample_a"]])
saveRDS(4:6, files[["sample_b"]])
store <- "pipeline"
```

## `hpc_initialise()`

This writes `{store}.R`, saves the input vector, and registers two mapped targets:

- `input_list` — the named input vector
- `sample_names` — the names of that vector

With no `computing_resources`, the pipeline is sequential.

```{r initialise}
pipe <- files |>
  hpc_initialise(store = store)
```

```{r initialise-script, comment = ""}
show_script(store)
```

## `is_target()`

Arguments to later steps are taken as literal values unless you wrap the name of an upstream target:

```{r is-target}
is_target("input_list")
```

That symbol is what `{targets}` wires as a dependency.

## `hpc_iterate()`

Map a function over every input. `file = "input_list" |> is_target()` means “use the `input_list` target, one element per branch”.

```{r iterate-read}
pipe <- pipe |>
  hpc_iterate(
    target_output = "data",
    user_function = readRDS |> quote(),
    file = "input_list" |> is_target()
  )
```

```{r iterate-read-script, comment = ""}
show_script(store)
```

A second iterate maps over the result of the first:

```{r iterate-length}
pipe <- pipe |>
  hpc_iterate(
    target_output = "n",
    user_function = length |> quote(),
    x = "data" |> is_target()
  )
```

```{r iterate-length-script, comment = ""}
show_script(store)
```

## `hpc_single()`

One target, not mapped over samples. Use this for file-tracking targets, scalars, or anything that should run once.

```{r single}
pipe <- pipe |>
  hpc_single(
    target_output = "n_samples",
    user_function = length |> quote(),
    x = "sample_names" |> is_target()
  )
```

```{r single-script, comment = ""}
show_script(store)
```

`hpc_initialise()` itself uses `hpc_single()` for the `*_file` targets (`format = "file"`).

## `hpc_merge()`

Collect every branch of an iterated target into one object.

```{r merge}
pipe <- pipe |>
  hpc_merge(
    target_output = "n_total",
    user_function = (function(lengths) sum(unlist(lengths))) |> quote(),
    lengths = "n" |> is_target()
  )
```

```{r merge-script, comment = ""}
show_script(store)
```

## `hpc_report()`

Appends a `tarchetypes::tar_quarto_raw()` factory. Rendering happens later, when you evaluate the pipeline; the grammar only writes the target.

```{r report-file}
writeLines(
  c(
    "---",
    "title: Example report",
    "format: html",
    "params:",
    "  n_total: 0",
    "  n_samples: 0",
    "---",
    "",
    "Samples: `r params$n_samples`. Values: `r params$n_total`."
  ),
  "example-report.qmd"
)
```

```{r report}
pipe <- pipe |>
  hpc_report(
    target_output = "report",
    rmd_path = "example-report.qmd",
    n_total = "n_total" |> is_target(),
    n_samples = "n_samples" |> is_target()
  )
```

```{r report-script, comment = ""}
show_script(store)
```

## `hpc_evaluate()`

Closes the script with a final `target_list` expression and calls `targets::tar_make()`. Printing a `tidytargets` object does the same thing.

```{r evaluate, eval = FALSE}
pipe |> hpc_evaluate()
```

The last two lines of the script after evaluate are:

``` r
target_list
```

Do not print the pipeline object unless you intend to run it.

## Tiers

`tier` splits inputs across named controllers. The factory then emits one target per tier (`data_1`, `data_2`, …) instead of a single mapped target.

```{r tiers}
files3 <- c(
  sample_a = "a.rds",
  sample_b = "b.rds",
  sample_c = "c.rds"
)
saveRDS(7:9, files3[["sample_c"]])

tiered_store <- "tiered_pipeline"

tiered <- files3 |>
  hpc_initialise(
    store = tiered_store,
    tier = c(1, 1, 2)
  ) |>
  hpc_iterate(
    target_output = "data",
    user_function = readRDS |> quote(),
    file = "input_list" |> is_target()
  )
```

```{r tiers-script, comment = ""}
show_script(tiered_store)
```

Controller names must match the `tier` labels when you pass a `{crew}` controller group:

```{r crew, eval = FALSE}
computing_resources <- crew::crew_controller_group(
  crew::crew_controller_local(name = "1", workers = 4),
  crew::crew_controller_local(name = "2", workers = 10)
)

files3 |>
  hpc_initialise(
    store = "tiered_pipeline",
    tier = c(1, 1, 2),
    computing_resources = computing_resources
  )
```

tidytargets does not import a compute backend. `computing_resources` is passed through to `targets::tar_option_set(controller = )`.

## Custom functions from a file

`hpc_iterate()`, `hpc_single()`, and `hpc_merge()` accept `user_function_source_path`. The path is written as `source(...)` in the targets script so workers can load helpers that are not in a package.

```{r source-path, eval = FALSE}
pipe |>
  hpc_iterate(
    target_output = "transformed",
    user_function = my_transform |> quote(),
    user_function_source_path = "helpers.R",
    x = "data" |> is_target()
  )
```
