---
title: "Building blocks of tidytargets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building blocks of tidytargets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.output = "r"
)
library(tidytargets)

work <- tempfile("tidytargets-vignette-")
dir.create(work)
knitr::opts_knit$set(root.dir = work)

show_script <- function(store) {
  path <- paste0(store, ".R")
  cat("## ", path, "\n\n", sep = "")
  writeLines(readLines(path))
}
```

tidytargets is a tidy, pipe-friendly grammar for `{targets}`. `tt_initialise()` constructs a `tidytargets` object; the other building blocks are S3 methods on that class. Each method is a [target factory](https://books.ropensci.org/targets/static.html#target-factories): it appends `tar_target` definitions to a pipeline script. You compose the pipeline with `|>`; `{targets}` runs the dependency graph.

The composition is **lazy**. `tt_initialise()`, `tt_iterate()`, `tt_single()`, `tt_merge()`, and `tt_report()` only write factories into `{store}.R`. User functions are not called until `tt_evaluate()` (or printing the `tidytargets` object), which closes the script and runs `targets::tar_make()`.

Pipelines are **incremental**. Add a step, change a parameter, or grow the input vector and `{targets}` skips up-to-date targets, re-running only what the new graph requires. That is why this vignette can show the script after each building block without executing the pipeline.

This vignette walks through every building block and prints the targets script after each step.

## Building blocks

| Function | Role |
| --- | --- |
| `tt_initialise()` | Start a pipeline: write the script header, register inputs |
| `tt_iterate()` | Map a function over inputs |
| `tt_single()` | Add one non-iterated target |
| `tt_merge()` | Combine iterated results into one object |
| `tt_report()` | Render a Quarto / R Markdown report |
| `tt_evaluate()` | Close the script and run `tar_make()` |
| `tt_factory()` | Internal factory used by iterate / single / merge |

`tt_factory()` is not meant to be called directly. `tt_iterate()`, `tt_single()`, and `tt_merge()` all append `target_list |> target_append(tt_factory(...))` to the script.

## Toy inputs

`tt_initialise()` takes a named vector of inputs, typically file paths. Here two RDS files stand in for samples.

```{r inputs}
files <- c(
  sample_a = "a.rds",
  sample_b = "b.rds"
)
saveRDS(1:3, files[["sample_a"]])
saveRDS(4:6, files[["sample_b"]])
store <- "pipeline"
```

## `tt_initialise()`

This writes `{store}.R`, saves the input vector, and registers two mapped targets:

- `input_list` — the named input vector
- `sample_names` — the names of that vector

With no `computing_resources`, the pipeline is sequential.

```{r initialise}
pipe <- files |>
  tt_initialise(store = store)
```

```{r initialise-script, comment = ""}
show_script(store)
```

## `tt_iterate()`

Map a function over every input. Wrap an upstream target name with `is_target()` so the argument is a dependency, not a literal string.

```{r iterate-read}
pipe <- pipe |>
  tt_iterate(
    target_output = "data",
    user_function = readRDS |> quote(),
    file = "input_list" |> is_target()
  )
```

```{r iterate-read-script, comment = ""}
show_script(store)
```

A second iterate maps over the result of the first:

```{r iterate-length}
pipe <- pipe |>
  tt_iterate(
    target_output = "n",
    user_function = length |> quote(),
    x = "data" |> is_target()
  )
```

```{r iterate-length-script, comment = ""}
show_script(store)
```

## `tt_single()`

One target, not mapped over samples. Use this for file-tracking targets, scalars, or anything that should run once.

```{r single}
pipe <- pipe |>
  tt_single(
    target_output = "n_samples",
    user_function = length |> quote(),
    x = "sample_names" |> is_target()
  )
```

```{r single-script, comment = ""}
show_script(store)
```

`tt_initialise()` itself uses `tt_single()` for the `*_file` targets (`format = "file"`).

## `tt_merge()`

Collect every branch of an iterated target into one object.

```{r merge}
pipe <- pipe |>
  tt_merge(
    target_output = "n_total",
    user_function = (function(lengths) sum(unlist(lengths))) |> quote(),
    lengths = "n" |> is_target()
  )
```

```{r merge-script, comment = ""}
show_script(store)
```

## `tt_report()`

Appends a `tarchetypes::tar_quarto_raw()` factory. Rendering happens later, when you evaluate the pipeline; the grammar only writes the target.

```{r report-file}
writeLines(
  c(
    "---",
    "title: Example report",
    "format: html",
    "params:",
    "  n_total: 0",
    "  n_samples: 0",
    "---",
    "",
    "Samples: `r params$n_samples`. Values: `r params$n_total`."
  ),
  "example-report.qmd"
)
```

```{r report}
pipe <- pipe |>
  tt_report(
    target_output = "report",
    rmd_path = "example-report.qmd",
    n_total = "n_total" |> is_target(),
    n_samples = "n_samples" |> is_target()
  )
```

```{r report-script, comment = ""}
show_script(store)
```

## `tt_evaluate()`

This is the only eager step. It closes the script with a final `target_list` expression and calls `targets::tar_make()`. Printing a `tidytargets` object does the same thing. Until then the pipeline is just a script on disk.

```{r evaluate, eval = FALSE}
pipe |> tt_evaluate()
```

The last two lines of the script after evaluate are:

``` r
target_list
```

Do not print the pipeline object unless you intend to run it.

## Custom functions from a file

`tt_iterate()`, `tt_single()`, and `tt_merge()` accept `user_function_source_path`. The path is written as `source(...)` in the targets script so workers can load helpers that are not in a package.

```{r source-path, eval = FALSE}
pipe |>
  tt_iterate(
    target_output = "transformed",
    user_function = my_transform |> quote(),
    user_function_source_path = "helpers.R",
    x = "data" |> is_target()
  )
```
