---
title: "Building blocks of tidytargets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building blocks of tidytargets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.output = "r"
)
library(tidytargets)
library(dplyr)

work <- tempfile("tidytargets-vignette-")
dir.create(work)
knitr::opts_knit$set(root.dir = work)
```

tidytargets is a tidy, pipe-friendly grammar for `{targets}`. 

`tt_initialise()` constructs a `tidytargets` object. Each method is a [target factory](https://books.ropensci.org/targets/static.html#target-factories): it appends `tar_target` definitions to a pipeline script. You compose the pipeline with `|>`; `{targets}` runs the dependency graph.

The composition is **lazy**. `tt_initialise()`, `tt_data()`, `tt_data_list()`, `tt_single()`, `tt_iterate()`, `tt_split()`, `tt_merge()`, and `tt_report()` only write factories into `{store}.R`. User functions are not called until `tt_evaluate()` (or evaluating the object; e.g., printing it in the console), which closes the script and runs `targets::tar_make()`.
Assigning the object does not print it, so an interactive session then says the pipeline is ready to be evaluated, rather than appearing to do nothing.

Pipelines are **incremental**. Add a step, change a parameter, or grow the input list and `{targets}` skips up-to-date targets, re-running only what the new graph requires. That is why this vignette can show the script after each building block without executing the pipeline.

This vignette walks through every building block and prints the targets script after each step with `show_targets_script()`.

## Building blocks

### Workflow

| Function | Role |
| --- | --- |
| `tt_initialise()` | Start a pipeline: write the script header, optional mapped inputs |
| `tt_evaluate()` | Close the script and run `tar_make()` |
| `tt_read()` | Read the full stored value of a named target |

### Data

| Function | Role |
| --- | --- |
| `tt_data()` | Snapshot a session object onto the store as one target |
| `tt_data_list()` | Snapshot a list onto the store as mapped units |
| `tt_merge()` | Combine iterated results into one object |
| `tt_split()` | Turn a pipeline list into mapped units |

### Commands

| Function | Role |
| --- | --- |
| `tt_iterate()` | Map a function over inputs |
| `tt_single()` | Add one non-iterated target |

`tt_factory()` is internal and used by the command functions above. `tt_single()`, `tt_iterate()`, `tt_split()`, and `tt_merge()` all append `target_list <- target_list |> target_append(tt_factory(...))` to the script.

### Reporting and debug

| Function | Role |
| --- | --- |
| `tt_report()` | Render a Quarto / R Markdown report |
| `tt_explore()` | Return one stored instance of a named target |
| `tt_metadata()` | Get or set free-form metadata on the pipeline object |

## Toy inputs

`tt_initialise()` can be called with no inputs (script header and store only). A named list of in-memory objects, one element per unit of iteration (e.g. sample), registers mapped targets. `tt_data()` adds a session object later without splitting it; `tt_data_list()` adds a list as mapped units.

```{r inputs}
inputs <- list(
  sample_a = 1:3,
  sample_b = 4:6
)
```

## `tt_initialise()`

This writes `{store}.R` and saves configuration. If you passed mapped inputs, it also registers:

- `input_list` — the named input vector or list (override with `target_output`)
- `sample_names` — the names of that vector or list

With no `computing_resources`, the pipeline is sequential. With no `store`, a unique `./tidytargets-<HASH>` directory is created. With no inputs, start from `tt_initialise()` and add objects with `tt_data()`.

Packages currently attached in your session (`.packages()`) are recorded at `tt_initialise()` as names on `tar_option_set(packages = )` so workers can `library()` them. `tt_initialise()` messages that list so you can check the same packages are available on HPC. Objects in the global environment are not copied; bring those in with `tt_data()` or `user_function_source_path`. Pass `packages` to override the snapshot. Attach libraries before initialise (or pass `packages`) if a later command needs them.

```{r initialise}
pipe <- inputs |>
  tt_initialise()
```

```{r initialise-script, comment = ""}
show_targets_script(pipe)
```

## `tt_data()`

Write a local object onto the store as a single target. Unlike `tt_initialise()`, the value is not split into iteration units.

```{r data}
extra <- list(label = "demo")
pipe <- pipe |>
  tt_data(extra)
```

```{r data-script, comment = ""}
show_targets_script(pipe)
```

## `tt_data_list()`

Same snapshot, but each list element is a mapped unit. Use this for a parameter grid split into rows (`group_split(row_number())`). Name an inline expression with `<-`, or pass `target_output`.

```{r data-list}
grid <- expand.grid(alpha = c(0, 1), lambda = c(0.1, 1))
pipe <- pipe |>
  tt_data_list(settings <- grid |> group_split(row_number()))
```

```{r data-list-script, comment = ""}
show_targets_script(pipe)
```

## `tt_single()`

This command adds an operation on a whole target (not iterative/grouped).

```{r summarise}
pipe <- pipe |>
  tt_single(n_samples <- length(sample_names))
```

```{r summarise-script, comment = ""}
show_targets_script(pipe)
```

Mapped lists from `tt_initialise()` and `tt_data_list()` are registered as stem list targets (no `map()` in the script) and marked as units for later `tt_iterate()`. File-tracking targets are stems too.

## `tt_iterate()`

Map a function over every input. Write `name <- expr` as you would a `tar_target()` command (`target_output = "name"` still works).

```{r iterate-read}
pipe <- pipe |>
  tt_iterate(data <- input_list * 2)
```

```{r iterate-read-script, comment = ""}
show_targets_script(pipe)
```

A second iterate maps over the result of the first:

```{r iterate-length}
pipe <- pipe |>
  tt_iterate(n <- length(data))
```

```{r iterate-length-script, comment = ""}
show_targets_script(pipe)
```

With two mapped inputs, `tt_iterate()` uses `map()` by default. `{targets}` errors at make if the lengths differ; pass `pattern = "cross"` for a product of branches. A constant (including a length-1 list) belongs in `tt_data()`, not `tt_data_list()`.

## `tt_evaluate()`

```{r evaluate}
pipe |> tt_evaluate()
```

## `tt_explore()`

Return one stored instance of a target so you can inspect it or pipe it onward. A heading names the target (and which instance, for collections). Mapped targets load a single branch rather than every sample.

```{r tt-explore}
pipe |> tt_explore(data)
```

## `tt_read()`

Read the full stored value of a named target (every branch of a mapped target). Compare with `tt_explore()` when you only need one instance.

```{r tt-read}
pipe |> tt_read(data)
```

## `tt_merge()`

Collect every branch of an iterated target into one object.

```{r merge}
pipe <- pipe |>
  tt_merge(n_total <- sum(unlist(n)))
```

```{r merge-script, comment = ""}
show_targets_script(pipe)
```

## `tt_split()`

Inverse of `tt_merge()`: a stem command that returns a list is marked as mapped units. Later `tt_iterate()` steps that mention it branch over the elements. Unlike `tt_data_list()`, the split runs in the pipeline.

```{r split}
grid_split <- expand.grid(alpha = c(0, 1), lambda = c(0.1, 1))
split_pipe <- tt_initialise() |>
  tt_data(grid_split) |>
  tt_split(settings <- grid_split |> group_split(row_number()))
```

```{r split-script, comment = ""}
show_targets_script(split_pipe)
```

## `tt_report()`

Appends a `tarchetypes::tar_quarto_raw()` factory. Rendering happens later, when you evaluate the pipeline; the grammar only writes the target.

```{r report-file}
writeLines(
  c(
    "---",
    "title: Example report",
    "format: html",
    "params:",
    "  n_total: 0",
    "  n_samples: 0",
    "---",
    "",
    "Samples: `r params$n_samples`. Values: `r params$n_total`."
  ),
  "example-report.qmd"
)
```

```{r report}
pipe <- pipe |>
  tt_report(
    target_output = "report",
    rmd_path = "example-report.qmd",
    params = list(n_total = n_total, n_samples = n_samples)
  )
```

```{r report-script, comment = ""}
show_targets_script(pipe)
```

## Custom functions from a file

`tt_iterate()`, `tt_single()`, `tt_split()`, and `tt_merge()` accept `user_function_source_path`. The path is written as `source(...)` in the targets script so workers can load helpers that are not in a package.

```{r source-path, eval = FALSE}
pipe |>
  tt_iterate(
    transformed <- my_transform(data),
    user_function_source_path = "helpers.R"
  )
```
